---
title: "STATS 295 Linear Setup Code"
output: html_document
---


### This is the Direct Method ATE estimation considering a predictor X, treatment A, and response Y


```{r}
set.seed(123)

sims <- 1000  # Number of simulations
n <- 1000     # Number of observations

# True ATE
ATE <- 0.5

# Direct Method (DM) results
D.M <- numeric(sims)

# Simulate data and compute estimates for each method
for (j in 1:sims) {
  X <- runif(n, 0, 1)  # Simulate X as uniform random variable between 0 and 1
  A <- ifelse(X > 0.6, 1, 0)  # Treatment assignment: A = 1 if X > 0.6 else A = 0
  epsilon <- rnorm(n, 0, 1)  # Error term
  
  # Response variable Y as given: Y = 0.5 * X + A * X + epsilon
  Y <- 0.5 * X + A * X + epsilon  
  
  # Direct Method (DM) Estimator
  fit <- lm(Y ~ A + X + A:X)  # Fit the linear model

  # Extract coefficients from the model
  coef <- summary(fit)$coefficients
  
  # Compute potential outcomes for A = 1 and A = 0
  con_mean1 <- coef[1, 1] + coef[2, 1] + (coef[3, 1] + coef[4, 1]) * X  # E[Y | A = 1]
  con_mean0 <- coef[1, 1] + coef[3, 1] * X  # E[Y | A = 0]
  
  # Compute ATE for this simulation using the Direct Method
  D.M[j] <- mean(con_mean1 - con_mean0)  # ATE using Direct Method
}

# Bias calculations
D.M_bias <- mean(ATE - D.M)  # Bias for Direct Method

# Variance calculations
D.M_var <- var(D.M)  # Variance for Direct Method

# Results summary for Direct Method only
results <- data.frame(
  Bias = D.M_bias,
  Variance = D.M_var
)

# Print the results
print(results)
```


### This is the Direct Method ATE estimation considering a predictor X, treatment A, response Y, and confounder Z1. Here we DO NOT adjust for the confounder Z1


```{r}
set.seed(123)

sims <- 1000  # Number of simulations
n <- 1000     # Number of observations

# True ATE
ATE <- 0.5

# Direct Method (DM) results
D.M <- numeric(sims)

# Simulate data and compute estimates for each method
for (j in 1:sims) {
  Z1 <- rnorm(n)

# Simulate X, now dependent on Z1 (e.g., X = 0.5*Z1 + noise)
X <- 0.5 * Z1 + runif(n)

  A <- ifelse(X > 0.6, 1, 0)  # Treatment assignment: A = 1 if X > 0.6 else A = 0
  epsilon <- rnorm(n, 0, 1)  # Error term
  
  # Response variable Y as given: Y = 0.5 * X + A * X + epsilon
  
  Y <- 0.5 * X + A * X + 0.5 * Z1 + epsilon
  
  # Direct Method (DM) Estimator
  fit <- lm(Y ~ A + X + A:X)  # Fit the linear model

  # Extract coefficients from the model
  coef <- summary(fit)$coefficients
  
  # Compute potential outcomes for A = 1 and A = 0
  con_mean1 <- coef[1, 1] + coef[2, 1] + (coef[3, 1] + coef[4, 1]) * X  # E[Y | A = 1]
  con_mean0 <- coef[1, 1] + coef[3, 1] * X  # E[Y | A = 0]
  
  # Compute ATE for this simulation using the Direct Method
  D.M[j] <- mean(con_mean1 - con_mean0)  # ATE using Direct Method
}

# Bias calculations
D.M_bias <- mean(ATE - D.M)  # Bias for Direct Method

# Variance calculations
D.M_var <- var(D.M)  # Variance for Direct Method

# Results summary for Direct Method only
results <- data.frame(
  Bias = D.M_bias,
  Variance = D.M_var
)

# Print the results
print(results)
```



### This is the Direct Method ATE estimation considering a predictor X, treatment A, response Y, and confounder Z1. Here we DO adjust for the confounder Z1

```{r}
set.seed(123)

sims <- 1000  # Number of simulations
n <- 1000     # Number of observations

# True ATE
ATE <- 0.5

# Direct Method (DM) results
D.M <- numeric(sims)

# Simulate data and compute estimates for each method
for (j in 1:sims) {
  Z1 <- rnorm(n)

# Simulate X, now dependent on Z1 (e.g., X = 0.5*Z1 + noise)
X <- 0.5 * Z1 + runif(n)

  A <- ifelse(X > 0.6, 1, 0)  # Treatment assignment: A = 1 if X > 0.6 else A = 0
  epsilon <- rnorm(n, 0, 1)  # Error term
  
  # Response variable Y as given: Y = 0.5 * X + A * X + epsilon
  
  Y <- 0.5 * X + A * X + 0.5 * Z1 + epsilon
  
  # Direct Method (DM) Estimator
  fit <- lm(Y ~ A + X + A:X + Z1)

  # Extract coefficients from the model
  coef <- summary(fit)$coefficients
  
  # Compute potential outcomes for A = 1 and A = 0 (adjusted for Z1)
  con_mean1 <- coef[1, 1] + coef[2, 1] + (coef[3, 1] + coef[5, 1]) * X + coef[4, 1] * Z1  # E[Y | A = 1]
  con_mean0 <- coef[1, 1] + coef[3, 1] * X + coef[4, 1] * Z1  # E[Y | A = 0]
  
  # Compute ATE for this simulation using the Direct Method
  D.M[j] <- mean(con_mean1 - con_mean0)  # ATE using Direct Method
}

# Bias calculations
D.M_bias <- mean(ATE - D.M)  # Bias for Direct Method

# Variance calculations
D.M_var <- var(D.M)  # Variance for Direct Method

# Results summary for Direct Method only
results <- data.frame(
  Bias = D.M_bias,
  Variance = D.M_var
)

# Print the results
print(results)
```



### This is the Direct Method ATE estimation considering a predictor X, treatment A, response Y, and mediator variable Z2. Here we DO NOT adjust for the mediator variable Z2


```{r}
set.seed(123)

sims <- 1000  # Number of simulations
n <- 1000     # Number of observations

# True ATE
ATE <- 0.5

# Direct Method (DM) results
D.M <- numeric(sims)

# Simulate data and compute estimates for each method
for (j in 1:sims) {
  X <- runif(n, 0, 1)  # Simulate X as uniform random variable between 0 and 1
  A <- ifelse(X > 0.6, 1, 0)  # Treatment assignment: A = 1 if X > 0.6 else A = 0
  epsilon_M <- rnorm(n)  # Error term for mediator
Z2 <- 0.5 * X +epsilon_M 
  
  # Response variable Y as given: Y = 0.5 * X + A * X + epsilon
epsilon <- rnorm(n, 0, 1)  # Error term
  Y <- 0.5 * Z2 + 0.5 * X + A * X + epsilon  
  
  # Direct Method (DM) Estimator
  fit <- lm(Y ~ A + X + A:X)  # Fit the linear model

  # Extract coefficients from the model
  coef <- summary(fit)$coefficients
  
  # Compute potential outcomes for A = 1 and A = 0
  con_mean1 <- coef[1, 1] + coef[2, 1] + (coef[3, 1] + coef[4, 1]) * X  # E[Y | A = 1]
  con_mean0 <- coef[1, 1] + coef[3, 1] * X  # E[Y | A = 0]
  
  # Compute ATE for this simulation using the Direct Method
  D.M[j] <- mean(con_mean1 - con_mean0)  # ATE using Direct Method
}

# Bias calculations
D.M_bias <- mean(ATE - D.M)  # Bias for Direct Method

# Variance calculations
D.M_var <- var(D.M)  # Variance for Direct Method

# Results summary for Direct Method only
results <- data.frame(
  Bias = D.M_bias,
  Variance = D.M_var
)

# Print the results
print(results)
```

### This is the Direct Method ATE estimation considering a predictor X, treatment A, response Y, and mediator variable Z2. Here we DO adjust for the mediator variable Z2


```{r}
set.seed(123)

sims <- 1000  # Number of simulations
n <- 1000     # Number of observations

# True ATE
ATE <- 0.5

# Direct Method (DM) results
D.M <- numeric(sims)

# Simulate data and compute estimates for each method
for (j in 1:sims) {
  X <- runif(n, 0, 1)  # Simulate X as uniform random variable between 0 and 1
  A <- ifelse(X > 0.6, 1, 0)  # Treatment assignment: A = 1 if X > 0.6 else A = 0
  epsilon_M <- rnorm(n)  # Error term for mediator
M <- 0.5 * X +epsilon_M 
  
  # Response variable Y as given: Y = 0.5 * X + A * X + epsilon
epsilon <- rnorm(n, 0, 1)  # Error term
  Y <- 0.5 * M + 0.5 * X + A * X + epsilon  
  
  # Direct Method (DM) Estimator
  fit <- lm(Y ~ A + X + A:X + M)  # Fit the linear model

  # Extract coefficients from the model
  coef <- summary(fit)$coefficients
  
  # Compute potential outcomes for A = 1 and A = 0
  con_mean1 <- coef[1, 1] + coef[2, 1] + (coef[3, 1] + coef[5, 1]) * X + coef[4,1] * M # E[Y | A = 1]
  con_mean0 <- coef[1, 1] + coef[3, 1] * X + coef[4,1] * M  # E[Y | A = 0]
  
  # Compute ATE for this simulation using the Direct Method
  D.M[j] <- mean(con_mean1 - con_mean0)  # ATE using Direct Method
}

# Bias calculations
D.M_bias <- mean(ATE - D.M)  # Bias for Direct Method

# Variance calculations
D.M_var <- var(D.M)  # Variance for Direct Method

# Results summary for Direct Method only
results <- data.frame(
  Bias = D.M_bias,
  Variance = D.M_var
)

# Print the results
print(results)
```





### This is the Direct Method ATE estimation considering a predictor X, treatment A, response Y, and precision variable Z3. Here we DO NOT adjust for the precision variable Z3


```{r}
set.seed(123)

sims <- 1000  # Number of simulations
n <- 1000   # Number of observations

# True ATE
ATE <- 0.5

# Direct Method (DM) results
D.M <- numeric(sims)

# Simulate data and compute estimates for each method
for (j in 1:sims) {
  X <- runif(n, 0, 1)  # Simulate X as uniform random variable between 0 and 1
  Z3 <- rnorm(n,0,1)
  A <- ifelse(X > 0.6, 1, 0)  # Treatment assignment: A = 1 if X > 0.6 else A = 0

  epsilon <- rnorm(n, 0, 1)  # Error term
  
  
  # Response variable Y as given: Y = 0.5 * X + A * X + epsilon
  Y <- 0.5 * X + A * X + 0.5* Z3 + epsilon  
  
  # Direct Method (DM) Estimator
  fit <- lm(Y ~ A + X + A:X)  # Fit the linear model

  # Extract coefficients from the model
  coef <- summary(fit)$coefficients
  
  # Compute potential outcomes for A = 1 and A = 0
  con_mean1 <- coef[1, 1] + coef[2, 1] + (coef[3, 1] + coef[4, 1]) * X  # E[Y | A = 1]
  con_mean0 <- coef[1, 1] + coef[3, 1] * X  # E[Y | A = 0]
  
  # Compute ATE for this simulation using the Direct Method
  D.M[j] <- mean(con_mean1 - con_mean0)  # ATE using Direct Method
}

# Bias calculations
D.M_bias <- (mean(D.M) - 0.5)  # Bias for Direct Method

# Variance calculations
D.M_var <- var(D.M)  # Variance for Direct Method

# Results summary for Direct Method only
results <- data.frame(
  Bias = D.M_bias,
  Variance = D.M_var
)

# Print the results
print(results)
```


### This is the Direct Method ATE estimation considering a predictor X, treatment A, response Y, and precision variable Z3. Here we DO adjust for the precision variable Z3


```{r}
set.seed(123456)

sims <- 1000  # Number of simulations
n <- 1000   # Number of observations

# True ATE
ATE <- 0.5

# Direct Method (DM) results
D.M <- numeric(sims)

# Simulate data and compute estimates for each method
for (j in 1:sims) {
  X <- runif(n, 0, 1)  # Simulate X as uniform random variable between 0 and 1
  A <- ifelse(X > 0.6, 1, 0)  # Treatment assignment: A = 1 if X > 0.6 else A = 0
  P <- rnorm(n)
  epsilon <- rnorm(n, 0, 1)  # Error term
  
  
  # Response variable Y as given: Y = 0.5 * X + A * X + epsilon
  Y <- 0.5 * X + A * X + 0.5 * P + epsilon  
  
  # Direct Method (DM) Estimator
  fit <- lm(Y ~ A + X + A:X+P)  # Fit the linear model

  # Extract coefficients from the model
  coef <- summary(fit)$coefficients
  
  # Compute potential outcomes for A = 1 and A = 0
  con_mean1 <- coef[1, 1] + coef[2, 1] + (coef[3, 1] + coef[5, 1]) * X + coef[4, 1] * P  # E[Y | A = 1]
  con_mean0 <- coef[1, 1] + coef[3, 1] * X + coef[4, 1] * P  # E[Y | A = 0]
  
  # Compute ATE for this simulation using the Direct Method
  D.M[j] <- mean(con_mean1 - con_mean0)  # ATE using Direct Method
}

# Bias calculations
D.M_bias <- mean(ATE - D.M)  # Bias for Direct Method

# Variance calculations
D.M_var <- var(D.M)  # Variance for Direct Method

# Results summary for Direct Method only
results <- data.frame(
  Bias = D.M_bias,
  Variance = D.M_var
)

# Print the results
print(results)
```




### This is the Direct Method ATE estimation considering a predictor X, treatment A, response Y, and nuisance variable Z4. Here we DO NOT adjust for the nuisance variable Z4

```{r}
set.seed(123)

sims <- 1000  # Number of simulations
n <- 1000     # Number of observations

# True ATE
ATE <- 0.5

# Direct Method (DM) results
D.M <- numeric(sims)

# Simulate data and compute estimates for each method
for (j in 1:sims) {
  X <- runif(n, 0, 1)  # Simulate X as uniform random variable between 0 and 1
  A <- ifelse(X > 0.6, 1, 0)  # Treatment assignment: A = 1 if X > 0.6 else A = 0
  
  epsilon_M <- rnorm(n)  # Error term for mediator
Z4 <- 0.5* X + rnorm(n)
  
  
  # Response variable Y as given: Y = 0.5 * X + A * X + epsilon
epsilon <- rnorm(n, 0, 1)  # Error term
  Y <-  0.5 * X + A * X + epsilon 
  
  # Direct Method (DM) Estimator
  fit <- lm(Y ~ A + X + A:X)  # Fit the linear model

  # Extract coefficients from the model
  coef <- summary(fit)$coefficients
  
  # Compute potential outcomes for A = 1 and A = 0
  con_mean1 <- coef[1, 1] + coef[2, 1] + (coef[3, 1] + coef[4, 1]) * X  # E[Y | A = 1]
  con_mean0 <- coef[1, 1] + coef[3, 1] * X  # E[Y | A = 0]
  
  # Compute ATE for this simulation using the Direct Method
  D.M[j] <- mean(con_mean1 - con_mean0)  # ATE using Direct Method
}

# Bias calculations
D.M_bias <- mean(ATE - D.M)  # Bias for Direct Method

# Variance calculations
D.M_var <- var(D.M)  # Variance for Direct Method

# Results summary for Direct Method only
results <- data.frame(
  Bias = D.M_bias,
  Variance = D.M_var
)

# Print the results
print(results)
```



### This is the Direct Method ATE estimation considering a predictor X, treatment A, response Y, and nuisance variable Z4. Here we DO NOT adjust for the nuisance variable Z4

```{r}
set.seed(123)

sims <- 1000  # Number of simulations
n <- 1000     # Number of observations

# True ATE
ATE <- 0.5

# Direct Method (DM) results
D.M <- numeric(sims)

# Simulate data and compute estimates for each method
for (j in 1:sims) {
  X <- runif(n, 0, 1)  # Simulate X as uniform random variable between 0 and 1
  A <- ifelse(X > 0.6, 1, 0)  # Treatment assignment: A = 1 if X > 0.6 else A = 0
  
  epsilon_M <- rnorm(n)  # Error term for mediator
Z4 <- 0.5* X + rnorm(n)
  
  
  # Response variable Y as given: Y = 0.5 * X + A * X + epsilon
epsilon <- rnorm(n, 0, 1)  # Error term
  Y <-  0.5 * X + A * X + epsilon 
  
  # Direct Method (DM) Estimator
  fit <- lm(Y ~ A + X + A:X + Z4)  # Fit the linear model

  # Extract coefficients from the model
  coef <- summary(fit)$coefficients
  
  # Compute potential outcomes for A = 1 and A = 0
  con_mean1 <- coef[1, 1] + coef[2, 1] + (coef[3, 1] + coef[5, 1]) * X + coef[4,1]*M # E[Y | A = 1]
  con_mean0 <- coef[1, 1] + coef[3, 1] * X + coef[4,1]*M  # E[Y | A = 0]
  
  # Compute ATE for this simulation using the Direct Method
  D.M[j] <- mean(con_mean1 - con_mean0)  # ATE using Direct Method
}

# Bias calculations
D.M_bias <- mean(ATE - D.M)  # Bias for Direct Method

# Variance calculations
D.M_var <- var(D.M)  # Variance for Direct Method

# Results summary for Direct Method only
results <- data.frame(
  Bias = D.M_bias,
  Variance = D.M_var
)

# Print the results
print(results)
```

